---
title: "UG Anger-Anxiety data"
author: "G. Marchetti"
date: "5 15 2024"
format: 
  html:
    embed-resources: true   
  pdf:
    documentclass: scrartcl
    number-sections: false
    colorlinks: true
    include-in-header: 
      text: | 
        \usepackage[default]{fontsetup} 
---

```{=tex}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\setmathfont{NewCMMath-Regular.otf}
\newcommand{\ci}{\hspace{0.1em}\perp\hspace{-0.95em}\perp\, }
\renewcommand{\b}[1]{\boldsymbol{#1}}
```
```{r}
#| include: false
#| message: false
source("~/Documents/R_packages/cat_regression_chains/RCG/funs_RCG.R")
library("bnlearn") 
library("igraph")
library("ggm")
library("gRbase")
library("mvtnorm")
```

# Data 

The sample size is $N = 684$. The data are in **ggm**, 
but the names are wrong and do not correspond to Cox and Wermuth book Table 3.1. 
```{r}
data("anger")
nam <- c('Y','X','V','U')
dimnames(anger) <- list(nam,nam)
anger
```
Correlations (lower triangle) an partial correlations
(upper triangle)
```{r}
round(correlations(anger),2)
```

Notice that all the marginal correlations are large
while two partial correlations are quite small.

Test the hypotheses
$$
Y \ci U \mid XV, \quad \text { and } \quad 
X \ci V \mid YU.
$$

```{r}
N <- 684
P <- parcor(anger)
dev1 <- -N * log(1 - P[1,4]^2)
dev1
dev2 <- -N * log(1 - P[2,3]^2)
dev2 
```

Both test are separately not sgnificant

## Fitted UG model

```{r}
#| fig-width: 3
#| fig-height: 3  
G <- UG(~ X*U + V*U + Y*V + X*Y)
drawGraph(G)
out <- fitConGraph(G, anger, 684)
out
```

Single edge deviance for edge $Y \sim V$
```{r}
#| fig-width: 3
#| fig-height: 3 
G1 <-  UG(~ X*U + V*U + Y*V + X*Y + Y*U)
drawGraph(G1)
out1 <- fitConGraph(G1, anger, 684)
out1
```