---
title: "D-separation in DAGs"
author: "G. Marchetti"
date: "May 13ยง 2024"
format: 
  html:
    embed-resources: true
  pdf:
    documentclass: scrartcl
    number-sections: false
    colorlinks: true
    include-in-header: 
      text: | 
        \usepackage[default]{fontsetup} 
---

```{=tex}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\setmathfont{NewCMMath-Regular.otf}
\newcommand{\ci}{\hspace{0.1em}\perp\hspace{-0.95em}\perp\, }
\renewcommand{\b}[1]{\boldsymbol{#1}}
```
```{r}
#| include: false
#| message: false

library("bnlearn") 
library("ggm")
library("gRbase")
library("igraph")
```

# The Chest Clinic Example

Form Lauritzen and Spiegelhalter (1988)

> Shortness-of-breath (dyspnoea) may be due to tuberculosis, lung cancer or bronchitis, or none of them, or more than one of them. A recent visit to Asia increases the chances of tu- berculosis, while smoking is known to be a risk factor for both lung cancer and bronchitis. The results of a single chest X-ray do not discriminate between lung cancer and tuberculo- sis, as neither does the presence or absence of dyspnoea.

The associate DAG is the following

```{r}
 g <- DAG( asia~ asia, tub ~  asia, smoke ~ smoke, lung ~  smoke, bronc ~ smoke, either ~ lung + tub, xray ~  either, dysp ~ bronc + either)
co <- structure(c(92, 64, 95, 76, 43, 49, 8, 9, 
                  80, 71, 26, 45, 20, 45, 65, 35), dim = c(8L, 2L))
drawGraph(g, coor = co)
```

## D-separation 

```{r}
dSep(g, "tub", "smoke", c("dysp","xray"))
```

```{r}
dSep(g, "tub", "lung", "smoke")
```