---
title: "Log-linear models and conditional independence"
author: "G. Marchetti"
date: "May 13 2024"
format: 
  html:
    embed-resources: true
  pdf:
    documentclass: scrartcl
    number-sections: false
    colorlinks: true
    include-in-header: 
      text: | 
        \usepackage[default]{fontsetup} 
---

```{=tex}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\setmathfont{NewCMMath-Regular.otf}
\newcommand{\ci}{\hspace{0.1em}\perp\hspace{-0.95em}\perp\, }
\renewcommand{\b}[1]{\boldsymbol{#1}}
```
```{r}
#| include: false
#| message: false
source("~/Documents/R_packages/cat_regression_chains/RCG/funs_RCG.R")
library("bnlearn") 
library("ggm")
library("gRbase")
library("igraph")
```


# Log-linear model and factorization

A joint probability vector for the model of conditional independence: $X_2$ independent of $X_3$ given $X_1$. 

```{r}
u1 <-  0.2
u2 <- -0.2
u3 <-  1.2
u12 <- 0.8
u13 <- 0.5
u23 <- 0
u123 <-0
lam <- c(0, u1, u2, u12, u3, u13, 0, 0)

L <- matrix(c(1,1,0,1), 2, 2)
M <- L  %x% L %x% L
p <- exp(M %*% lam)
p <- p/sum(p)
p
```
The contingency table
```{r}
X <- expand.grid(X1 = c(0,1), X2 = c(0,1), X3 = c(0,1), stringsAsFactors = TRUE)
data3 <- data.frame(X, p)
data3
```
```{r}
ftable(X1 + X2 ~ X3, xtabs(p ~. , data3))
```
The conditional odds-ratio are both 1
```{r}
(0.05480844 * 0.14898478)/ (0.18197043 * 0.04487335)
```
```{r}
(0.06694318 * 0.24563438)/(0.13480701 * 0.12197842)
```

# An example

Some old data concerning breast cancer reported by Morrison n (1973). The three factors are

- $X_1$ diagnostic center
- $X_2$ nuclear grade 
- $X_3$ survival after three years 


Read the data 

```{r}
Freq <- c(35, 42, 59, 77, 47, 26, 112, 76)
df_bc <- data.frame(expand.grid(X1 = c("Boston", "Glamorgan"), X2 = c("malignant", "benign"), X3 = c("died", "survived"), stringsAsFactors = TRUE), Freq)
df_bc
```


Fit a saturated model 

```{r}
m_sat  <-glm(Freq ~ X1 * X2 * X3, family = poisson, data = df_bc)
m_sat
```

Fit a log-linear model

```{r}
m_ci <- glm(Freq ~ X1 * X2 + X1 * X3, family = poisson, data = df_bc)
m_ci
```

The likelihood ratio test is $G^2_2 = 4.072$ that is not significant. 

```{r}
anova(m_ci, m_sat, test = "Chisq")
```




## Fitted counts

```{r}
m <- fitted(m_ci)
data_fit <- cbind(df_bc[, 1:3], m = m)
data_fit
m[1] * m[7]/(m[3]*m[5])
m[2] * m[8]/(m[4]*m[6])
```
# Using a significance test for the same conditional independence

Use the package **bnlearn**  to test conditional independence

```{r}
library(readr)
data_bc<- read_rds("data_bc.rds")

ci.test("X2", "X3", "X1", data = data_bc)
```
