---
title: "UG Reinis data"
author: "G. Marchetti"
date: "5 15 2024"
format: 
  html:
    embed-resources: true
  pdf:
    documentclass: scrartcl
    number-sections: false
    colorlinks: true
    include-in-header: 
      text: | 
        \usepackage[default]{fontsetup} 
---

```{=tex}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\setmathfont{NewCMMath-Regular.otf}
\newcommand{\ci}{\hspace{0.1em}\perp\hspace{-0.95em}\perp\, }
\renewcommand{\b}[1]{\boldsymbol{#1}}
```
```{r}
#| include: false
#| message: false
source("~/Documents/R_packages/cat_regression_chains/RCG/funs_RCG.R")
library("bnlearn") 
library("igraph")
library("ggm")
library("gRbase")
library("mvtnorm")
```

# Reinis Data 

Data collected at the beginning of a 15 year follow-up study of probable risk factors for coronary thrombosis. Data are from all men employed in a car factory ($N= 1841$).



A table with 6 discrete variables. A: smoking, B: strenous mental work, C: strenuous physical work, D: systolic blood pressure, E: ratio of lipoproteins, F: Family anamnesis of coronary heart disease.



```{r}
library("readr")
data_reinis <- read_rds("data_reinis.rds")
colnames(data_reinis) <- c('A', 'B', 'C', 'D', 'E', 'F')
tab_reinis <- table(data_reinis)
head(as.data.frame(tab_reinis))
```

## Test of conditional independence

A conditional independence $ F \ci D \mid BECA$ is expected.
Let's test this hypothesis

```{r}
ci.test("F", "D", c("B", "E", "C", "A"), data = data_reinis)
```
The statistic `mi` is the same of the deviance. Apparently 
the conditional independence is not rejected.

The previous CI is implied for example by 
the following graph 

```{r}
#| fig-width: 4
#| fig-height: 2
G <- UG(~ F*C*B + B*E*C*A + E*A*D)
drawGraph(G, layout = layout_nicely)
```

## Fit the graph  

```{r}
m_red <- glm(Freq ~ F*C*B + B*E*C*A + E*A*D, 
             family = poisson, 
             data = as.data.frame(tab_reinis))
summary(m_red)
```

## The model shows some further possible reductions

The strongest terms correspond to `CB`, `CE`, `AD` and `CBA`. Therefore a suitable graph should be 
```{r}
G0 <- UG(~ F + C*B*A + C*E + A*D)
drawGraph(G0, layout = layout_nicely)
```
Notice that surprisingly ther is a complete independence 
of family anamnesis of CHD from the other variables.

Fitting this graphical model we get

```{r}
m0 <- glm(Freq ~  F + C*B*A + C*E + A*D,
             family = poisson, 
             data = as.data.frame(tab_reinis))
summary(m0)
```
The deviance however indicates a quite poor agreement with the hypothesis so this model is rejected.


